Basic regression in Gaussian processes

### 2019-10-28 manuscript/appendix/appendix.Rmd
greta, tensorflow GP, python, etc

```{r}
sim_data %>% group_by(reps) %>%
  sample_n(5e2) %>%
  ggplot(aes(t, x, group=reps)) + 
  geom_line(alpha=0.05) +
  geom_line(data = det, color = "blue") + 
  geom_line(data = mean, color = "red")
```

@@

arima_forecast <- sim_data %>% 
  filter(reps == lows[[1]]) %>% 
  select(t,x) %>%
  bind_cols(y)
```

```{r arima_figure, message=FALSE, warning=FALSE}
train <- arima_forecast %>% filter(t < 2000) 
test <- arima_forecast %>% filter(t > 2000) 
train %>%
  ggplot(aes(t)) + 
  geom_line(aes(y=x)) +
  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
  xlim(0,3000) + ylab("x") + theme_minimal()

@@

# a <- 2.3
r <- p$r
Q <- p$Q
K <- p$K
H <- p$H
sigma <- p$sigma

a <- uniform(0, 10)
r <- uniform(0, 2 * p$r)
#Q <- uniform(0, 10)
K <- uniform(0, 2*p$K)
H <- uniform(0,  2*p$H)
sigma <- uniform(0, 2 * p$sigma)

@@

p$x0 <- x0
p$N <- 1e3
true_forecast <- 
  future_map_dfr(1:100, 
                 function(reps) sim(may, p), 
                 .id = "reps"
                 ) %>% 
  mutate(t = t+2000) %>% 
  bind_rows(training_data) %>% 
  mutate(set = "true")

predicted_forecast <- 
  posterior_sims %>% 
    mutate(t = t+2000) %>%
    bind_rows(training_data) %>% 
    mutate(set = "predicted")
```

```{r}
bind_rows(predicted_forecast, true_forecast) %>% 
#predicted_forecast %>%
  ggplot(aes(t,x)) +
  geom_line(aes(group = interaction(reps, set), col = set), alpha = .05) + 
  geom_line(data = training_data) + 
  facet_wrap(~set) + 
  scale_color_solarized()
```

>>>>

```{r}
write_csv(sim_data, "../../data/ghost_sims.csv.gz")
sim_data %>% group_by(reps) %>%
  sample_n(5e2) %>%
  ggplot(aes(t, x, group=reps)) + 
  geom_line(alpha=0.05) +
  geom_line(data = det, color = "blue") + 
  geom_line(data = mean, color = "red")
```

@@

arima_forecast <- sim_data %>% 
  filter(reps == lows[[1]]) %>% 
  select(t,x) %>%
  bind_cols(y)

write_csv(arima_forecast, "../../data/arima_forecast.csv.gz")
```

```{r arima_figure, message=FALSE, warning=FALSE}

train <- arima_forecast %>% filter(t < 2000) 
test <- arima_forecast %>% filter(t > 2000) 
train %>%
  ggplot(aes(t)) + 
  geom_line(aes(y=x)) +
  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), 
              alpha = 0.4, fill = "red", data = test)  + 
  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
  xlim(0,3000) + ylab("x") + theme_minimal()

@@

a <- uniform(0, 10)
r <- uniform(0, 4 * p$r)
Q <- uniform(0, 4 * p$Q)
K <- uniform(0, 4 * p$K)
H <- uniform(0,  4 * p$H)
sigma <- uniform(0, 4 * p$sigma)

@@

p$x0 <- x0
p$N <- 1e3
true_forecast <- 
  future_map_dfr(1:100, 
                 function(reps) sim(may, p), 
                 .id = "reps"
                 ) %>% 
  mutate(t = t+2000) %>% 
  bind_rows(training_data) %>% 
  mutate(set = "true")
```

```{r}
predicted_forecast <- 
  posterior_sims %>% 
    mutate(t = t+2000) %>%
    bind_rows(training_data) %>% 
    mutate(set = "predicted")
```

```{r}
model_forecast <- bind_rows(predicted_forecast, true_forecast)
write_csv(model_forecast, "../../data/model_forecast.csv.gz")

model_forecast %>% 
#predicted_forecast %>%
  ggplot(aes(t,x)) +
  geom_line(aes(group = interaction(reps, set), col = set), alpha = .1) + 
  geom_line(data = training_data) + 
  facet_wrap(~set) + 
  scale_color_solarized()
```

### 2019-10-28 notebook/gp-manual-vs-greta.Rmd
greta, tensorflow GP, python, etc

```{r setup, message = FALSE, warning = FALSE}

require(MASS)
require(reshape2)
require(ggplot2)

@@

ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=gp, 
              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
              fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```

@@

# hyperparameters
rbf_var <- 10 # lognormal(0, 1)
rbf_len <- 0.5 #lognormal(0, .01)
obs_sd <- .01 #lognormal(0, .01)

@@

# likelihood -- not clear how this influences f_plot?
distribution(y) <- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_plot)
```

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, warmup = 0)
```
```

>>>>

```{r setup, message = FALSE, warning = FALSE}
library(MASS)
library(tidyverse)

@@

ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=gp, 
              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
              fill="grey80") + # Var
  geom_line(aes(group=variable), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) + #REPLICATES
  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```

@@

# hyperparameters
rbf_var <- lognormal(0, 1)
rbf_len <- lognormal(0, 1)
obs_sd <- lognormal(0, 1)

@@

# likelihood -- not clear how this influences f_plot?
distribution(y)<- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_predict)
```

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot, rbf_len, rbf_var, obs_sd)
draws <- mcmc(m, chains = 1, one_by_one = TRUE)
```

```{r}
df <- draws[[1]] %>% as_tibble(.name_repair = "universal") 
hyperpars <- df %>% dplyr::select(rbf_len, obs_sd, rbf_var)  
posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)

hyperpars %>% colMeans()
#hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
```

```{r}
posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
  ggplot(aes(x,value)) + 
  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
  geom_point(data = obs, aes(x,y))
```

```{r}
```

### 2019-10-28 notebook/greta-gp.Rmd
greta, tensorflow GP, python, etc

```{r}
arima_forecast <- readr::read_csv("../data/arima_forecast.csv.gz")
train <- arima_forecast  %>% filter(t %in% seq(0, 2000, by = 5))

@@

y <- c(0,x_t1 - x_t)
x <- c(0, x_t)
x_plot <- seq(0, 2, length.out = 200)
```

```{r}

```{r}
# simulate toy data
n<- 10
x <- runif(n, 0, 10)
y <- sin(x) + rnorm(n, 0, 0.5)
x_plot <- seq(-1, 14, length.out = 200)
```

```{r}

# hyperparameters
rbf_var <- 10 # lognormal(0, 1)
rbf_len <- 0.5 #lognormal(0, .01)
obs_sd <- .01 #lognormal(0, .01)

@@

# likelihood -- not clear how this influences f_plot?
distribution(y) <- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_plot)
```

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, warmup = 0)
```

```{r}
# plot 200 posterior samples
plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))

#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
for (i in 1:200) {
  lines(draws[[1]][i, ] ~ x_plot,
        lwd = 2,
        col = rgb(0.7, 0.1, 0.4, 0.1))  
}
points(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))

>>>>

## Model Definition

```{r}
p <- list(r = .05, K = 200, Q = 5, H = 38, sigma = .002, a=2.3, N = 3e3, x0 = 6)
```

```{r}
growth <- function(x, p) x * p$r * (1 - x / p$K)
consumption <- function(x, p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
may <- function(x, p) x + growth(x,p) - consumption(x,p)
```

## Model simulations

```{r}
# Generic simulator routine
sim <- function(f, p){
  x <- numeric(p$N)
  x[1] <- p$x0
  dBt <- numeric(p$N)
  if(p$sigma > 0) dBt <- rnorm(p$N, 0, p$sigma)
  for(t in 1:(p$N-1)){
    x[t+1] <- max(f(x[t], p) + x[t] * dBt[t], 0)
  }
  data.frame(t = 1:p$N, x)
}
```

Deterministic simulations:

```{r}
p_det <- p
p_det$sigma <- 0
det <- sim(may,p_det) 
det_diffs <- det %>% mutate(y = lead(x)-x, case = "det")
det_diffs <- det %>% mutate(y = c(diff(x),NA), case = "det")

ex <- sim(may,p) 
ex %>% mutate(y = lead(x)-x, case = "stoch")  %>% bind_rows(det_diffs) %>%
  ggplot(aes(x,y, col=case)) + geom_point()
```

```{r}
ghost <- readr::read_csv("../data/ghost_sims.csv.gz")
wide <- ghost %>% pivot_wider(names_from = reps, values_from = x) %>% select(-t)
n <- dim(wide)[1]
x_t1 <- wide[-1,]
x_t <- wide[-n,] 
y <- x_t1 - x_t
Y <- y %>% apply(1, mean, na.rm=TRUE)
X <- x_t %>% apply(1, mean, na.rm=TRUE)

tibble(x=X,y=Y, case="mean") %>% bind_rows(det_diffs) %>% ggplot(aes(x,y, col=case)) + geom_point()

ghost %>% group_by(reps) %>%  mutate(y = lead(x)-x) %>% 
  ggplot(aes(x,y)) + geom_point(size=.1, alpha=0.005)
```

```{r}
arima_forecast <- readr::read_csv("../data/arima_forecast.csv.gz")
train <- arima_forecast  #%>% filter(t %in% seq(0, 2000, by = 5))

@@

y <- c(0,x_t1 - x_t)
x <- c(0, x_t)
x_predict <- seq(0, 2, length.out = 200)

tibble(x,y) %>% ggplot(aes(x,y)) + geom_point()
```

----

```{r}
# kernel & GP
y <- diff(ex$x)
x <- ex$x[-n]
ggplot(data.frame(x,y), aes(x,y)) + geom_point()

# hyperparameters
rbf_var <- lognormal(0, 1)
rbf_len <- lognormal(0, 10)
obs_sd <- lognormal(0, .1)

@@

# likelihood -- not clear how this influences f_plot?
distribution(y)<- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_predict)
```

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot, rbf_len, rbf_var, obs_sd)
draws <- mcmc(m, chains = 1, one_by_one = TRUE)
```

```{r}
df <- draws[[1]] %>% as_tibble(.name_repair = "universal") 
hyperpars <- df %>% dplyr::select(rbf_len, obs_sd, rbf_var)  
posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)

hyperpars %>% colMeans()
#hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
```

```{r}
posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
  ggplot(aes(x,value)) + 
  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
  geom_point(data = obs, aes(x,y))
```

### 2019-10-28 notebook/python-reticulate-gp.Rmd
greta, tensorflow GP, python, etc

```{python}
import numpy as np
import tensorflow as tf
import tensorflow_probability as tfp
```

```{python}
tfd = tfp.distributions
psd_kernels = tfp.positive_semidefinite_kernels
```

```{python}
num_points = 100
# Index points should be a collection (100, here) of feature vectors. In this
# example, we're using 1-d vectors, so we just need to reshape the output from
# np.linspace, to give a shape of (100, 1).
index_points = np.expand_dims(np.linspace(-1., 1., num_points), -1)

# Define a kernel with default parameters.
kernel = psd_kernels.ExponentiatedQuadratic()
```

```{python}
gp = tfd.GaussianProcess(kernel, index_points)

samples = gp.sample(10)
# ==> 10 independently drawn, joint samples at `index_points`

noisy_gp = tfd.GaussianProcess(
    kernel=kernel,
    index_points=index_points,
    observation_noise_variance=.05)
noisy_samples = noisy_gp.sample(10)

noisy_samples
```

>>>>

```{r message=FALSE}
library(tidyverse)
library(reticulate)
library(tensorflow)
np <- import("numpy")
tf <- import("tensorflow")
tfp <- import("tensorflow_probability")
sess <- tf$Session()

```

```{r}
tfd <- tfp$distributions
psd_kernels <- tfp$positive_semidefinite_kernels
```

## Draws from Prior

```{r}
num_points = 100L
# Index points should be a collection (100, here) of feature vectors. In this
# example, we're using 1-d vectors, so we just need to reshape the output from
# np.linspace, to give a shape of (100, 1).
index_points = np$expand_dims(np$linspace(-1., 1., num_points), -1L)

var <- tf$constant(1.0, dtype = tf$float64)
len <-  tf$constant(0.1, dtype = tf$float64)
obs <-  tf$constant(0.05, dtype = tf$float64)

# Define a kernel with default parameters.
kernel = psd_kernels$ExponentiatedQuadratic(var, len)## Radial basis function
  #psd_kernels$ExponentiatedQuadratic()
```

```{r}
gp = tfd$GaussianProcess(kernel, index_points)

samples = gp$sample(10L)
# ==> 10 independently drawn, joint samples at `index_points`
```

Or with observation error:

```{r}
noisy_gp = tfd$GaussianProcess(
    kernel=kernel,
    index_points=index_points,
    observation_noise_variance=obs)

noisy_samples = noisy_gp$sample(10L)
```

Evaluate and extract results

```{r}
out <- sess$run(samples) 
```

```{r}
out %>% 
  t() %>% 
  as_tibble %>% 
  mutate(x = seq(-1, 1, length.out = 100) ) %>%
  pivot_longer(-x) %>%
  ggplot(aes(x, value, group = name)) + geom_line(col = "purple", alpha=0.4)
```

## Gaussian Process Regression

```{r}
# Generate noisy observations from a known function at some random points.
observation_noise_variance <- 0.01 #<-  tf$constant(obs, dtype = tf$float64)
len <-  tf$constant(0.1, dtype = tf$float64)
var <-  tf$constant(1, dtype = tf$float64)

f = function(x) sin(10 * x) * exp(-x ^ 2)
## Index points and observation_index_points is an array, with 1 col per feature! ## (N-dimensional X)
observation_index_points = runif(5L, -1, 1) %>% as.matrix(ncol = 1)
## Observations must be a numeric, not an array. (vector of values Y, not matrix of features)
observations = f(as.numeric(observation_index_points)) +
  runif(2L, 0., sqrt(observation_noise_variance))

index_points = seq(-1, 1, len=100L) %>% as.matrix(ncol = 1)

## This works too but no need to it in python, the above is fine!!
#index_points = np$linspace(-1., 1., 100L)  %>%  np$expand_dims(-1L)
#x = np$random$uniform(-1., 1., 5L) 
#observations = (np$sin(10*x) * np$exp(-x**2) + 
#                  np$random$normal(0., np$sqrt(observation_noise_variance)))
#observation_index_points <- x %>%  np$expand_dims(-1L)

kernel = psd_kernels$ExponentiatedQuadratic(var, len) # MaternFiveHalves()

gprm = tfd$GaussianProcessRegressionModel(
    kernel = kernel,
    index_points = index_points,
    observation_index_points=observation_index_points,
    observations = observations,
    observation_noise_variance = observation_noise_variance)

samples = gprm$sample(10L)
```

```{r}
out <- sess$run(samples) 
```

```{r}
ob <- data.frame(x = observation_index_points, y = observations, name = "data")

out %>% 
  t() %>% 
  as_tibble %>% 
  mutate(x = index_points) %>%
  pivot_longer(-x) %>%
  ggplot(aes(x, value, group = name)) + 
  geom_line(col = "purple", alpha=0.4) + 
  geom_point(data = ob, aes(x,y))
```

## ML

```{r}
# Define a kernel with trainable parameters. Note we transform the trainable
# variables to apply a positivity constraint.
amplitude = tf$exp(tf$Variable(1, dtype=np$float64), name='amplitude')
length_scale = tf$exp(tf$Variable(2, dtype=np$float64), name='length_scale')
kernel = psd_kernels$ExponentiatedQuadratic(amplitude, length_scale)

observation_noise_variance = tf$exp(
    tf$Variable(0.5, dtype=np$float64), name='observation_noise_variance')

# We'll use an unconditioned GP to train the kernel parameters.
gp = tfd$GaussianProcess(
    kernel=kernel,
    index_points= observation_index_points,
    observation_noise_variance=observation_noise_variance)
neg_log_likelihood = -gp$log_prob(observations)

optimizer = tf$train$AdamOptimizer(learning_rate=.05, beta1=.5, beta2=.99)
optimize = optimizer$minimize(neg_log_likelihood)

gprm = tfd$GaussianProcessRegressionModel(
    kernel=kernel,
    index_points=index_points,
    observation_index_points=observation_index_points,
    observations=observations,
    observation_noise_variance=observation_noise_variance)

ml_samples = gprm$sample(10L)

```

Note that we need to initialize variables!  
```{r}
init_op <- tf$global_variables_initializer()
with(tf$Session() %as% sess, {
   # Run the 'init' op
  sess$run(init_op)
  out <- sess$run(ml_samples)

  param_est <- sess$run(c(amp = amplitude, 
                  len = length_scale, 
                  obs = observation_noise_variance))
})
```

```{r}
x
ob <- data.frame(x = observation_index_points, y = observations, name = "data")

out %>% 
  t() %>% 
  as_tibble %>% 
  mutate(x = index_points) %>%
  pivot_longer(-x) %>%
  ggplot(aes(x, value, group = name)) + 
  geom_line(col = "purple", alpha=0.4) + 
  geom_point(data = ob, aes(x,y))
```

## MCMC

```{r}

joint_log_prob <- function(
    index_points, observations, amplitude, length_scale, noise_variance){
  # Hyperparameter Distributions
  rv_amplitude = tfd$LogNormal(tf$constant(0, dtype=np$float64),
                               tf$constant(1, dtype=np$float64))

  rv_length_scale = tfd$LogNormal(tf$constant(0, dtype=np$float64), 
                                  tf$constant(1, dtype=np$float64))

  rv_noise_variance = tfd$LogNormal(tf$constant(0, dtype=np$float64),
                                    tf$constant(1, dtype=np$float64))

  gp = tfd$GaussianProcess(
      kernel=psd_kernels$ExponentiatedQuadratic(amplitude, length_scale),
      index_points=index_points,
      observation_noise_variance=noise_variance)

  return (
      rv_amplitude$log_prob(amplitude) +
      rv_length_scale$log_prob(length_scale) +
      rv_noise_variance$log_prob(noise_variance) +
      gp$log_prob(observations)
  )
}
```

```{r}
scale <- tf$constant(1e-1, dtype=np$float64)
initial_chain_states = list(
    scale * tf$ones(1, dtype=np$float64, name='init_amplitude'),
    scale * tf$ones(1, dtype=np$float64, name='init_length_scale'),
    scale * tf$ones(1, dtype=np$float64, name='init_obs_noise_variance')
)

# Since HMC operates over unconstrained space, we need to transform the
# samples so they live in real-space.
unconstraining_bijectors = list(
    tfp$bijectors$Softplus(),
    tfp$bijectors$Softplus(),
    tfp$bijectors$Softplus()
)

unnormalized_log_posterior <- function(amplitude, length_scale, noise_variance){

   joint_log_prob(observation_index_points, 
                  observations, 
                  amplitude, 
                  length_scale,
                  noise_variance)

}

num_results = 200L

# list(
  #   list(
  #     amplitudes,
  #     length_scales,
  #     observation_noise_variances
  #     ),
  #   kernel_results
  # )
```

```{r}
out = tfp$mcmc$sample_chain(
  num_results = num_results,
  num_burnin_steps = 500L,
  num_steps_between_results = 3L,
  current_state = initial_chain_states,
  kernel = tfp$mcmc$TransformedTransitionKernel(
    inner_kernel = tfp$mcmc$HamiltonianMonteCarlo(
      target_log_prob_fn = unnormalized_log_posterior,
      step_size= tf$constant(.15, dtype = np$float64),
      num_leapfrog_steps = 3L
    ),
    bijector = unconstraining_bijectors
  )
)

# Now we can sample from the posterior predictive distribution at a new set
# of index points.
```
```{r}
#with(tf$Session() %as% sess){
#  out <- sess$run(out)
#}
```

```{r}
gprm = tfd$GaussianProcessRegressionModel(
    # Batch of `num_results` kernels parameterized by the MCMC samples.
    kernel=psd_kernels$ExponentiatedQuadratic(out[[0]][[1]], out[[0]][[2]]),
    index_points=index_points,
    observation_index_points=observation_index_points,
    observations=observations,
    observation_noise_variance=out[[0]][[3]])

## sample!
samples = gprm$sample()
```

```{r}
samples_ = sess$run(samples)

res = sess$run(c(kernel_results = out, samples = samples))

```

```{python}
  print("Acceptance rate: {}".format(
      np$mean(kernel_results_.inner_results.is_accepted)))

  # Plot posterior samples and their mean, target function, and observations.
  plt.plot(np.stack([index_points[:, 0]]*num_results).T,
           samples_.T,
           c='r',
           alpha=.01)
  plt.plot(index_points[:, 0], np.mean(samples_, axis=0), c='k')
  plt.plot(index_points[:, 0], f(index_points))
  plt.scatter(observation_index_points[:, 0], observations)
}
```
